Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-01-31 paper.tex
progress on DSL induction section

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} $\lambda \text{var}.$ body
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p:\alpha\to\cdots\to\beta \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if canUnify}(\tau,\beta) \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Let $e:\alpha_1\to\alpha_2\to\cdots\to \alpha_K\to\beta$. Unify $\tau$ with $\beta$.
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} $e\;a_1\; a_2\; \cdots\; a_K$
 %  \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[tb]
   \caption{Grammar Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $\log \probability[\mathcal{D}]\stackrel{+}{ = } -\lambda\sum_{p\in \mathcal{D}} \text{size}(p)$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ a subtree of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \IF{$\text{score}(\mathcal{D}') > \text{score}(\mathcal{D})$}
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ELSE
     \STATE\textbf{return} $\mathcal{D},\theta^*(\mathcal{D})$
     \ENDIF
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}
\section{Program Representation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.

>>>>

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} $\lambda \text{var}.$ body
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p: \tau' \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if }\tau\text{ can unify with yield}(\tau')) \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Let $e:\alpha_1\to\alpha_2\to\cdots\to \alpha_K\to\beta$. Unify $\tau$ with $\beta$.
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} $e\;a_1\; a_2\; \cdots\; a_K$
 \STATE
 \STATE \textbf{function} yield$(\tau)$:
 \STATE \textbf{Input:} type $\tau$
 \STATE \textbf{Output:} the return type of $\tau$
 \STATE \textbf{if} $\tau = \alpha\to \beta$ \textbf{then return }yield$(\beta)$ \textbf{else return }$\tau$
\end{algorithmic}
\end{algorithm}

@@

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at the programs in the frontiers and
generalize beyond them; 
not only so the DSL can better express the current solutions,
but also so that the DSL might expose new abstractions
which will later used to
discover even better programs.

Exact maximization of $J$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $J$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.

Closely related to Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
but context-sensitive

We define a prior distribution over DSLs which penalizes the sizes of the $\lambda$-calculus expressions in the DSL, and put a Dirichlet prior over the weight vector:
\begin{align*}
  \probability[\mathcal{D}]&\propto\exp\left(\lambda\sum_{p\in \mathcal{D}}\text{size}(p) \right)\\
  P(\theta|\mathcal{D})& = \text{Dir}(\theta|\alpha)
\end{align*}
where $\text{size}(p)$  measures the size of the syntax tree of program $p$,
$\lambda$ is a hyperparameter that acts as a regularizer on the size of the DSL,
and $\alpha$ is a concentration parameter controlling the smoothness of the prior over $\theta$.
Algorithm~\ref{grammarInductionAlgorithm} specifies the DSL induction algorithm.

Additionally, for each proposed $\mathcal{D}$ we have to reestimate $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar (PCFG: see~\cite{}),
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
$\mathcal{D}$
 is actually context-sensitive
due to the presence of variables in the programs and also due to the
polymorphic typing system.
In the Appendix we derive an efficient MAP estimator for $\theta$.

\begin{algorithm}[tb]
   \caption{DSL Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}

\begin{figure*}
  \begin{tabular}{lll}
    \toprule
    Domain&Example programs in frontiers&Example proposed subexpression\\\midrule
    Boolean circuits&
    \begin{tabular}{l}
      \texttt{(lambda x (nand x x))}\\
      \texttt{(lambda x (lambda y (nand x (nand y y))}
    \end{tabular}&
    \texttt{(nand z z)}\\\midrule
    String editing&
    \begin{tabular}{l}
      \texttt{(lambda s (+ ',' (index 0 (split ',' s))))}\\
      \texttt{(lambda s (index 0 (split ' ' s)))}
    \end{tabular}&
    \texttt{(index 0 (split z s))}\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm works by proposing subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (middle column), and can introduce new variables (\texttt{z} in the right column).}
  \label{fragmentExample}
  \end{figure*}
\section{Program Representation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.
